Ultrafast

### 2024-01-25 manuscript/manuscript.qmd
Revisions

affiliations:
  - id: eth
    name: "ETH Zürich"
    department: "Institute of Chemical and Bioengineering, Department of Chemistry and Applied Biosciences"
    address: "Vladimir-Prelog-Weg 1"
    postal-code: "CH-8093"
    city: "Zürich"
    country: "Switzerland"
  - id: chalmers
    name: "Chalmers University of Technology"
    department: "Department of Chemistry and Chemical Engineering"
    address: "Kemigården 4"
    postal-code: "SE-41258"
    city: "Gothenburg"
    country: "Sweden"
  - id: toronto-chem
    name: "University of Toronto"
    department: "Chemical Physics Theory Group, Department of Chemistry"
    address: "80 St. George St."
    postal-code: "M5S 3H6"
    city: "Toronto"
    state: "Ontario"
    country: "Canada"
  - id: toronto-cs
    name: "University of Toronto"
    department: "Department of Computer Science"
    address: "40 St. George St."
    postal-code: "M5S 2E4"
    city: "Toronto"
    state: "Ontario"
    country: "Canada"
  - id: rug
    name: "University of Groningen"
    department: "Stratingh Institute for Chemistry"
    address: "Nijenborgh 4"
    postal-code: "9747 AG"
    city: "Groningen"
    country: "The Netherlands"
  - id: toronto-chem-eng
    name: "University of Toronto"
    department: "Department of Chemical Engineering & Applied Chemistry"
    address: "200 College St."
    postal-code: "M5S 3E5"
    city: "Toronto"
    state: "Ontario"
    country: "Canada"
  - id: toronto-mat-eng
    name: "University of Toronto"
    department: "Department of Materials Science & Engineering"
    address: "184 College St."
    postal-code: "M5S 3E4"
    city: "Toronto"
    state: "Ontario"
    country: "Canada"    
  - id: vector
    name: "Vector Institute for Artificial Intelligence"
    address: "661 University Ave. Suite 710"
    postal-code: "M5G 1M1"
    city: "Toronto"
    state: "Ontario"
    country: "Canada" 
  - id: cifar
    name: "Lebovic Fellow"
    department: "Canadian Institute for Advanced Research (CIFAR)"
    postal-code: "M5G 1M1"
    city: "Toronto"
    state: "Ontario"
    country: "Canada" 
  - id: ac
    name: "Acceleration Consortium"
    department: "University of Toronto"
    city: "Toronto"
    state: "Ontario"
    country: "Canada"

title: "Ultrafast computational screening of molecules with inverted singlet-triplet energy gaps using the Pariser-Parr-Pople semi-empirical quantum chemistry method"
abstract: |
  Molecules with an inverted energy gap between their first singlet and triplet excited states have promising applications in the next generation of organic light-emitting diode (OLED) materials. Unfortunately, such molecules are rare and only a handful of examples are currently known. High-throughput virtual screening could assist in finding novel classes of these molecules, but current efforts are hampered by the high computational cost of the required quantum-chemical methods. We present a method based on the semi-empirical Pariser-Parr-Pople theory augmented by perturbation theory and show that it reproduces inverted gaps at a fraction of the cost of currently employed excited state calculations.. Our study paves the way for ultra-high-throughput virtual screening and inverse design to accelerate the discovery and development of this new generation of OLED materials.
bibliography: [references.bib] 
link-citations: true
date: last-modified
format:
  elsevier-pdf:
    cite-method: citeproc
    journal:
      name: "Journal of Physical Chemistry A"
      model: "1p"    
    number-sections: false
    keep-tex: true            
    include-in-header:
        - file: macros.tex    
csl: "acs.csl"
---

@@

tab_local_azaphenalenes = ""
tab_local_azulenes = ""
tab_global_all_gaps = ""
tab_global_scaffolds = ""
tab_garner_gaps = ""
```

@@

Molecules with an INVErted Singlet-Triplet energy gap (INVEST) are exceedingly rare.[@liOrganicMoleculesInverted2022] They are said to violate Hund's rule of maximum multiplicity as applied to the S~1~ and T~1~ states, which describes that the electronic state with the highest spin (in this case the triplet) should be lowest in energy. While a handful of examples of molecules violating Hund's rule in the excited state have been known since the 1970s and 1980s,[@kollmarViolationHundRule1978; @leupinLowlyingElectronicallyExcited1980; @kosekiViolationHundMultiplicity1985;@toyotaViolationHundRule1988;@toyotaViolationHundMultiplicity1986] it is only recently that they have received considerable attention for application in OLEDs.[@desilvaInvertedSingletTriplet2019;@polliceOrganicMoleculesInverted2021;@aizawaDelayedFluorescenceInverted2022] As they are rare, recent efforts have used high-throughput virtual screening (HTVS) with computational chemistry to identify new compounds with inverted gaps, focusing on expanding the local chemical space around specific scaffolds[@polliceOrganicMoleculesInverted2021;@sanz-rodrigoNegativeSingletTriplet2021;@ricciSingletTripletExcited2021;@sobolewskiAreHeptazineBasedOrganic2021;@aizawaDelayedFluorescenceInverted2022;@ricciEstablishingDesignStrategies2022;@tuckovaOriginInvertedSinglet2022] or scanning larger (combinatorially or experimentally derived) datasets to identify novel scaffolds.[@terenceblaskovitsSymmetryInducedSinglet2023;@omarIdentificationUnknownInverted2023;@garnerDoublebondDelocalizationNonalternant2023] These HTVS efforts have been successful at identifying several new INVEST candidates, some of which have also been synthesized and tested experimentally.[@aizawaDelayedFluorescenceInverted2022]

While this early success of HTVS is highly encouraging, it is hampered by the cost of the computational methods. Excitation energies are routinely calculated by time-dependent density functional theory (TD-DFT), which has become a workhorse of computational photophysics over the last decades.[@casidaProgressTimeDependentDensityFunctional2012] Unfortunately, it has been shown that TD-DFT fails to capture the inverted gap of INVEST compounds, as the method only considers single electronic excitations.[@desilvaInvertedSingletTriplet2019] Inclusion of at least double excitations are necessary to reproduce the inverted gaps, corresponding to methods such as double hybrid TD-DFT or excited state coupled-cluster methods such as second-order approximate coupled cluster singles and doubles (CC2) or equation of motion coupled cluster with single and double excitations (EOM-CCSD). In line with the results of benchmarking studies,[@ricciSingletTripletExcited2021;@sancho-garciaViolationHundRule2022] recent HTVS studies have used methods such as the complete active space self-consistent field (CASSCF) and CC2 for preliminary screening, while confirming inverted gaps with more expensive methods such as multistate complete active space second-order perturbation theory (MS-CASPT2) and EOM-CCSD. These methods are not only expensive, but in the case of the CAS methods, also require the choice of an orbital active space, something which is not routinely automatized.

@@

While the recent literature has focused on applying high-level *ab initio* and DFT methods to study INVEST compounds, we believe that some conceptual clarity has been lost in the process. Instead, we apply the simplest possible electronic structure method that can capture the physics of the problem. The Pariser-Parr-Pople (PPP) method is an extension of the semi-empirical Hückel molecular orbital theory for $\pi$-electron systems that includes electron correlation.[@pariserSemiEmpiricalTheory1953a;@pariserSemiEmpiricalTheory1953; @popleElectronInteractionUnsaturated1953] It originated in the 1950s, with new parametrizations being developed mainly in the 1960s and 1970s, and was used in the dye industry at least until the 1980s before the advent of more accurate ab initio methods.[@griffithsPracticalAspectsColour1982] The PPP method retains the conceptual clarity of Hückel theory, while at the same time including the electron correlation that is necessary to capture the inverted gap.

@@

according to recursion formulas given by Mulliken.[@mullikenFormulasNumericalTables1949] Thus, the four parameters $U_{r}$, $\gamma_{rr}$, $\gamma_{rs}$ and $\beta_{rs}$ are expressed completely in terms of the corresponding valence state ionization potentials $\mathrm{IP}_v$ and electron affinities $\mathrm{EA}_v$ as well as the completely empirical parameter $C$.

@@

$$
\beta_{rs}^\prime = \beta_{rs} \cos{\theta}
$${#eq-beta-rs}

@@

A more precise expression for $\DeltaE{ST}$ can be obtained by configuration interaction singles (CIS).[@beveridgeParametrizationSemiempiricalPi1971] While CIS adds some correlation for the excited states, it is necessary to add additional excitations beyonds singles to capture the inverted $\DeltaE{ST}$. Fortunately, the most important excitations have already been identified in the literature long ago. Singlet-triplet inversion occurs also for the ground state, where several violations of Hund’s rule are well known, for example for bond-equalized cyclobutadiene at the *D*~4h~ geometry. Borden and Davidson explained this effect in terms of dynamic spin polarization (DSP),[@bordenTheoreticalStudiesDiradicals1981] in which the electrons of a pair of disjoint non-bonding molecular orbitals (NBMOs, [@fig-dsp]a) experience stronger electron correlation effects in the singlet state as compared to the triplet state ([@fig-dsp]b). For a pedagogic introduction to the topic, the reader is referred to an excellent article by Karafiloglou.[@karafiloglouDoubleDynamicSpin1989] As outlined by Kollmar and Staemmler,[@kollmarViolationHundRule1978] as well as Malrieu and co-workers,[@benamorSpinPolarizationElectronic2020] an alternative view of the same phenomenon is through admixture of singly-excited configurations into the electronic wavefunction of the ground state ([@fig-dsp]c). The DSP effect for the S~1~ excited state can correspondingly be described by admixture of doubly excited configurations,[@kollmarViolationHundRule1978] as they are singly excited with respect to the S~1~ state, and doubly excited with respect to the S~0~ ground state ([@fig-dsp]d).

As first shown by Kollmar and Staemmler,[@kollmarViolationHundRule1978] and recently re-emphasized by Pernal and co-workers,[@drwalRoleSpinPolarization2023] the $\DeltaE{ST}$ can be approximated by a combination of two terms, the first one being the exchange interaction $2K$ (given by @eq-exchange) and a correction term $\DeltaE[DSP]{}$ due to the DSP, which can be approximated with perturbation theory:[@kollmarViolationHundRule1978]

![Dynamic spin polarization stabilizes the open-shell singlet state over the open-shell triplet. (a) Transformation of the canonical frontier molecular orbitals creates a set of disjoint non-bonding molecular orbitals for cyclobutadiene. Singlet and triplet occupations are shown. (b) Dynamic spin polarization preferentially stabilizes the singlet state of cyclobutadiene through additive spin polarization for the singlet and competitive spin polarization for the triplet. (c) An alternative description of the spin polarization phenomenon is via configuration interaction and admixture of excited configurations into the ground state. (d) In the same way as cyclobutadiene is stabilized by DSP in the singlet ground state, molecules with inverted singlet-triplet gaps in the excited state are also stabilized by DSP that can be described by admixture of excited configurations. For singly excited states, this means addition of doubly excited states.](../data/figures/dynamic_spin_polarization.pdf){#fig-dsp}

@@

The PPP method was implemented in the Python package [coulson]{.smallcaps}[@jornerCoulson] which is freely available on GitHub with a permissive open-source MIT license. Further details on [coulson]{.smallcaps} will be reported elsewhere. The PPP wavefunction was converged with the self-consistent field method using the naive variational principle. To speed up the calculations, we derived a spline interpolation of the overlap integrals,[@mullikenFormulasNumericalTables1949] which will be reported separately. All calculations have been performed on a MacBook Air laptop computer with an M2 processor. We further integrated [coulson]{.smallcaps} with [PySCF]{.smallcaps}[@sunRecentDevelopmentsPySCF2020] to provide alternative algorithms for the SCF convergence and CIS and found the results consistent within numerical accuracy. Oscillator strengths were calculated with the dipole length approximation.[@jorgensenTimedependentHartreeFockCalculations1970;@klessingerExcitedStatesPhotochemistry1995;@damourGroundExcitedStateDipole2023] Some compounds exhibit negative triplet excitation energies and in some case even negative singlet excitation energies, which indicates a restricted-unrestricted instability of the ground state wavefunction. We have here taken the pragmatic approach to ignore these issues, while we will highlight below some examples where it occurs and if it has any effect on the overall results. This approach can be partly justified as the reference data also does not include any stability analysis. The results should anyway be indicative of the gap between the lowest singlet and triplet states with open-shell character, regardless of if they are lower in energy than the closed-shell singlet.

We used four separate datasets for this study. The first is a set of azaphenalenes previously studied by some of us, comprising 256 substituted compounds, for which excitation energies were computed with (2nd-order) algebraic-diagrammatic construction, {{< var reference_level >}}/cc-pVDZ, and geometries were optimized with B3LYP/cc-pVDZ[@polliceOrganicMoleculesInverted2021] The second is a set substituted 138 azaazulenes with {{< var reference_level >}}/cc-pVDZ excitation energies and optimized with the B97-3c composite method.[@nigamArtificialDesignOrganic2023] The third is a set of 16 rationally designed scaffolds, which including substitutions amounts to {{< var global_n_total >}} unique compounds, optimized at the B97-3c level and with excitation energies at the {{< var reference_level >}}/cc-pVDZ level.[@polliceRationalDesignOrganic2023] The fourth is a set 315 non-alternant hydrocarbons including substitutions, divided into three subsets of size 76, 187 and 52, optimized at the $\omega$B97X-D/def2-TZVP level and with excitation energies at the CC2/aug-cc-pVTZ level.

@@

We first demonstrate that our method is capable of capturing the inverted $\DeltaE{ST}$ for some model compounds. As shown by Toyota and co-workers, pentalene at the ideal  $D_\mathrm{2h}$ geometry has an inverted gap, while relaxation to the minimum with $C_\mathrm{2h}$ symmetry leads to a normal gap.[@kosekiViolationHundMultiplicity1985] In our calculations, pentalene with equal bond lengths of 1.4 Å displays a small HOMO-LUMO overlap of {{< var pentalene_ideal_overlap >}}, leading to a small exchange interaction of only $2K=$ {{< var pentalene_ideal_2_exchange >}} kcal/mol (@fig-pentalene-ideal-orbitals). The dynamic spin polarization correction $\DeltaE[DSP]{SCF}$ of {{< var pentalene_ideal_dsp_scf >}} kcal/mol leads to a net predicted $\DeltaEST[DSP]{SCF}$ of {{< var pentalene_ideal_t1_s1_dsp_scf >}} kcal/mol. Adding additional correlation with CIS leads to a $\DeltaEST[DSP]{CIS}$ of {{< var pentalene_ideal_t1_s1_dsp_cis >}} kcal/mol. The simple perturbation theory model allows us to gain additional insight into the contributions to $\DeltaE[DSP]{SCF}$ as they correspond to single excitations from doubly occupied orbitals below the HOMO to unoccupied orbitals above the LUMO (@fig-pentalene-ideal-excitations). Three excitations, HOMO-3 → LUMO+2, HOMO-2 → LUMO+1 and HOMO-1 → LUMO+3 are identified as the main contributors to the spin polarization effect, and could potentially be tuned by substituent effects.

![Excitations and their contribution to $\DeltaE[DSP]{SCF}=$ {{< var pentalene_ideal_dsp_scf >}} kcal/mol for pentalene at the $D_\mathrm{2h}$ geometry and bond lengths of 1.4 Å.]({{< var fig_pentalene_ideal_excitations >}}.png){#fig-pentalene-ideal-excitations}

Using the relaxed $C_\mathrm{2h}$ geometry leads to dramatic changes in the frontier orbitals, with an increased HOMO-LUMO overlap of {{< var pentalene_overlap >}} and a sizeable exchange interaction of {{< var pentalene_2_exchange >}} kcal/mol (@fig-pentalene-orbitals). Consequently, the calculated gap is now normal at $\DeltaEST[DSP]{SCF}=$ {{< var pentalene_t1_s1_dsp_scf >}} kcal/mol, aggravated by a diminished $\DeltaE[DSP]{SCF}$ of only {{< var pentalene_dsp_scf >}} kcal/mol. A similar result is obtained with CIS: $\DeltaEST[DSP]{CIS}=$ {{< var pentalene_t1_s1_dsp_cis >}} kcal/mol. The reference value calculated with {{< var reference_level >}} is {{< var pentalene_t1_s1_ref >}} kcal/mol. The excitations contributing to $\DeltaE[DSP]{SCF}$ at the $D_\mathrm{2h}$ geometry have been significantly diminished, only partially alleviated by the addition of a minor stabilization from HOMO-2 → LUMO+3 (@fig-pentalene-excitations).

![Excitations and their contribution to $\DeltaE[DSP]{SCF}=$ {{< var pentalene_dsp_scf >}} kcal/mol for pentalene at the optimized $C_\mathrm{2h}$ geometry.]({{< var fig_pentalene_excitations >}}.png){#fig-pentalene-excitations}

>>>>

affiliations:
  - id: eth
    name: "ETH Zürich"
    department: "Institute of Chemical and Bioengineering, Department of Chemistry and Applied Biosciences"
    address: "Vladimir-Prelog-Weg 1"
    postal-code: "CH-8093"
    city: "Zürich"
    country: "Switzerland"
  - id: chalmers
    name: "Chalmers University of Technology"
    department: "Department of Chemistry and Chemical Engineering"
    address: "Kemigården 4"
    postal-code: "SE-41258"
    city: "Gothenburg"
    country: "Sweden"
  - id: toronto-chem
    name: "University of Toronto"
    department: "Chemical Physics Theory Group, Department of Chemistry"
    address: "80 St. George St."
    postal-code: "M5S 3H6"
    city: "Toronto"
    state: "Ontario"
    country: "Canada"
  - id: toronto-cs
    name: "University of Toronto"
    department: "Department of Computer Science"
    address: "40 St. George St."
    postal-code: "M5S 2E4"
    city: "Toronto"
    state: "Ontario"
    country: "Canada"
  - id: rug
    name: "University of Groningen"
    department: "Stratingh Institute for Chemistry"
    address: "Nijenborgh 4"
    postal-code: "9747 AG"
    city: "Groningen"
    country: "The Netherlands"
  - id: toronto-chem-eng
    name: "University of Toronto"
    department: "Department of Chemical Engineering & Applied Chemistry"
    address: "200 College St."
    postal-code: "M5S 3E5"
    city: "Toronto"
    state: "Ontario"
    country: "Canada"
  - id: toronto-mat-eng
    name: "University of Toronto"
    department: "Department of Materials Science & Engineering"
    address: "184 College St."
    postal-code: "M5S 3E4"
    city: "Toronto"
    state: "Ontario"
    country: "Canada"    
  - id: vector
    name: "Vector Institute for Artificial Intelligence"
    address: "661 University Ave. Suite 710"
    postal-code: "M5G 1M1"
    city: "Toronto"
    state: "Ontario"
    country: "Canada" 
  - id: cifar
    name: "Lebovic Fellow"
    department: "Canadian Institute for Advanced Research (CIFAR)"
    address: "661 University Ave."
    postal-code: "M5G 1M1"
    city: "Toronto"
    state: "Ontario"
    country: "Canada" 
  - id: ac
    name: "Acceleration Consortium"
    department: "University of Toronto"
    address: 700 University Ave.
    postal-code: "M5G 1Z5"
    city: "Toronto"
    state: "Ontario"
    country: "Canada"

title: "Ultrafast Computational Screening of Molecules with Inverted Singlet-triplet Energy Gaps Using the Pariser-Parr-Pople Semi-empirical Quantum Chemistry Method"
abstract: |
  Molecules with an inverted energy gap between their first singlet and triplet excited states have promising applications in the next generation of organic light-emitting diode (OLED) materials. Unfortunately, such molecules are rare and only a handful of examples are currently known. High-throughput virtual screening could assist in finding novel classes of these molecules, but current efforts are hampered by the high computational cost of the required quantum-chemical methods. We present a method based on the semi-empirical Pariser-Parr-Pople theory augmented by perturbation theory and show that it reproduces inverted gaps at a fraction of the cost of currently employed excited state calculations. Our study paves the way for ultra-high-throughput virtual screening and inverse design to accelerate the discovery and development of this new generation of OLED materials.

bibliography: [references.bib] 
link-citations: true
date: last-modified
format:
  elsevier-pdf:
    cite-method: citeproc
    journal:
      name: "Journal of Physical Chemistry A"
      model: "3p"    
    number-sections: false
    keep-tex: true
    include-in-header:
      - file: macros.tex    
      - file: packages.tex
csl: "acs.csl"   
---

@@

tab_local_azaphenalenes = ""
tab_local_azulenes = ""
tab_global_all_gaps = ""
tab_global_scaffolds = ""
tab_garner_gaps = ""
tab_parameters = ""
tab_opt = ""
```

@@

Molecules with an INVErted Singlet-Triplet energy gap (INVEST) are exceedingly rare.[@liOrganicMoleculesInverted2022] They \deleted{are said to} violate Hund's rule of maximum multiplicity as applied to the S~1~ and T~1~ states \added{of molecules}, which describes that the electronic state with the highest spin (in this case the triplet) should be lowest in energy. While a handful of examples of molecules violating Hund's rule in the excited state have been known since the 1970s and 1980s,[@kollmarViolationHundRule1978; @leupinLowlyingElectronicallyExcited1980; @kosekiViolationHundMultiplicity1985;@toyotaViolationHundRule1988;@toyotaViolationHundMultiplicity1986] it is only recently that they have received considerable attention for application in OLEDs.[@desilvaInvertedSingletTriplet2019;@polliceOrganicMoleculesInverted2021;@aizawaDelayedFluorescenceInverted2022] As they are rare, recent efforts have used high-throughput virtual screening (HTVS) with computational chemistry to identify new compounds with inverted gaps, focusing on expanding the local chemical space around specific scaffolds[@polliceOrganicMoleculesInverted2021;@sanz-rodrigoNegativeSingletTriplet2021;@ricciSingletTripletExcited2021;@sobolewskiAreHeptazineBasedOrganic2021;@aizawaDelayedFluorescenceInverted2022;@ricciEstablishingDesignStrategies2022;@tuckovaOriginInvertedSinglet2022] or scanning larger (combinatorially or experimentally derived) datasets to identify novel scaffolds.[@terenceblaskovitsSymmetryInducedSinglet2023;@omarIdentificationUnknownInverted2023a;@garnerDoublebondDelocalizationNonalternant2023a] These HTVS efforts have been successful at identifying several new INVEST candidates, some of which have also been synthesized and tested experimentally.[@aizawaDelayedFluorescenceInverted2022]

While this early success of HTVS is highly encouraging, it is hampered by the cost of the computational methods. Excitation energies are routinely calculated by time-dependent density functional theory (TD-DFT), which has become a workhorse of computational photophysics over the last decades.[@casidaProgressTimeDependentDensityFunctional2012] Unfortunately, it has been shown that TD-DFT fails to capture the inverted gap of INVEST compounds, as the method only considers single electronic excitations.[@desilvaInvertedSingletTriplet2019] Inclusion of at least double excitations are necessary to reproduce the inverted gaps, corresponding to methods such as double hybrid TD-DFT \added{(via perturbation theory)}[@grimmeDoublehybridDensityFunctional2007] or excited state coupled-cluster methods such as second-order approximate coupled cluster singles and doubles (CC2) or equation of motion coupled cluster with single and double excitations (EOM-CCSD). In line with the results of benchmarking studies,[@ricciSingletTripletExcited2021;@sancho-garciaViolationHundRule2022] recent HTVS studies have used methods such as the complete active space self-consistent field (CASSCF) and CC2 for preliminary screening, while confirming inverted gaps with more expensive methods such as multistate complete active space second-order perturbation theory (MS-CASPT2) and EOM-CCSD. These methods are not only expensive \added{(compared to TD-DFT)}, but in the case of the CAS methods, also require the choice of an orbital active space, something which is not routinely automatized \added{(although advances have been made)}.[@steinAutomatedSelectionActive2016;@sayfutyarovaAutomatedConstructionMolecular2017]

@@

While the recent literature has focused on applying high-level *ab initio* and DFT methods to study INVEST compounds, we believe that some conceptual clarity has been lost in the process. Instead, we apply the simplest possible electronic structure method that can capture the physics of the problem. The Pariser-Parr-Pople (PPP) method is an extension of the semi-empirical Hückel molecular orbital theory for $\pi$-electron systems that includes electron correlation.[@pariserSemiEmpiricalTheory1953a;@pariserSemiEmpiricalTheory1953; @popleElectronInteractionUnsaturated1953] It originated in the 1950s, with new parametrizations being developed mainly in the 1960s and 1970s, and was used in the dye industry at least until the 1980s before the advent of more accurate ab initio methods.[@griffithsPracticalAspectsColour1982] The PPP method retains the conceptual clarity of Hückel theory, while at the same time including the electron correlation that is necessary to capture the inverted gap. \added{For another recent application of PPP to INVEST compounds, see the work by Painelli and co-workers.}[@bedogniShiningLightInverted2023]

@@

according to recursion formulas given by Mulliken.[@mullikenFormulasNumericalTables1949] Thus, the four parameters $U_{r}$, $\gamma_{rr}$, $\gamma_{rs}$ and $\beta_{rs}$ are expressed completely in terms of the corresponding valence state ionization potentials $\mathrm{IP}_v$ and electron affinities $\mathrm{EA}_v$ as well as the completely empirical parameter $C$. \added{We have used the set of valence state ionization potentials and electron affinities from Beveridge and Hinze} ([@tbl-parameters]).[@beveridgeParametrizationSemiempiricalPi1971]

@@

$$
\beta_{rs}^\prime = \beta_{rs} \cos{\theta}
$${#eq-beta-rs-prime}

@@

A more precise expression for $\DeltaE{ST}$ can be obtained by configuration interaction singles (CIS).[@beveridgeParametrizationSemiempiricalPi1971] While CIS adds some correlation for the excited states, it is necessary to add additional excitations beyonds singles to capture the inverted $\DeltaE{ST}$. Fortunately, the most important excitations have already been identified in the literature long ago. Singlet-triplet inversion occurs also for the ground state, where several violations of Hund’s rule are well known, for example for bond-equalized cyclobutadiene at the *D*~4h~ geometry. Borden and Davidson explained this effect in terms of dynamic spin polarization (DSP),[@bordenTheoreticalStudiesDiradicals1981] in which the electrons of a pair of disjoint non-bonding molecular orbitals (NBMOs, [@fig-dsp]a) experience stronger electron correlation effects in the singlet state as compared to the triplet state ([@fig-dsp]b).[@albrightOrbitalInteractionsChemistry2013] For a pedagogic introduction to the topic, the reader is referred to an excellent article by Karafiloglou.[@karafiloglouDoubleDynamicSpin1989] As outlined by Kollmar and Staemmler,[@kollmarViolationHundRule1978] as well as Malrieu and co-workers,[@benamorSpinPolarizationElectronic2020; @davidImprovedEvaluationSpinpolarization2020] an alternative view of \replaced{spin polarization}{the same phenomenon} is through admixture of \deleted{singly-}excited configurations into the electronic wavefunction of the ground state ([@fig-dsp]c).[@chipmanSpinPolarizationModel1992;@engelsRecentDevelopmentsConfiguration1996;@coulaudAnalysisPhysicalContributions2012] \added{This is shown in }[@fig-dsp]\added{c for the case of static spin polarization in the allyl radical.} The DSP effect for the S~1~ excited state can correspondingly be described by admixture of doubly excited configurations,[@kollmarViolationHundRule1978] \deleted{as they are} singly excited with respect to the S~1~ state, and doubly excited with respect to the S~0~ ground state ([@fig-dsp]d).

As first shown by Kollmar and Staemmler,[@kollmarViolationHundRule1978] and recently re-emphasized by Pernal and co-workers,[@drwalRoleSpinPolarization2023a] the $\DeltaE{ST}$ can be approximated by a combination of two terms, the first one being the exchange interaction $2K$ (given by @eq-exchange) and a correction term $\DeltaE[DSP]{}$ due to the DSP, which can be approximated with perturbation theory:[@kollmarViolationHundRule1978]

![Dynamic spin polarization stabilizes the open-shell singlet state over the open-shell triplet. (a) Transformation of the canonical frontier molecular orbitals creates a set of disjoint non-bonding molecular orbitals for cyclobutadiene. Singlet and triplet occupations are shown. (b) Dynamic spin polarization preferentially stabilizes the singlet state of cyclobutadiene through additive spin polarization for the singlet and competitive spin polarization for the triplet. (c) An alternative description of the spin polarization phenomenon is via configuration interaction and admixture of excited configurations into the ground state. (d) In the same way as cyclobutadiene is stabilized by DSP in the singlet ground state, molecules with inverted singlet-triplet gaps in the excited state are also stabilized by DSP that can be described by admixture of excited configurations. For singly excited states, this means addition of doubly excited states. Panel b has been adapted from ref. [@karafiloglouDoubleDynamicSpin1989].](../data/figures/dynamic_spin_polarization.pdf){#fig-dsp}

@@

The PPP method was implemented in the Python package [coulson]{.smallcaps}[@jornerCoulson] which is freely available on GitHub with a permissive open-source MIT license. Further details on [coulson]{.smallcaps} will be reported elsewhere. The PPP wavefunction was converged with the self-consistent field method using the naive variational principle. \added{CIS calculations used the converged SCF wavefunction as reference and modelled vertical excitations based on the pre-computed geometries.} To \replaced{make}{speed up} the calculations \added{more robust}, we derived a spline interpolation of the overlap integrals,[@mullikenFormulasNumericalTables1949] which will be reported separately. \added{Evaluated on the azaphenalene dataset (vide infra), it shows good accuracy with $R^2=$} {{< var interpolation_all_r2_cis >}} \added{and RMSE =} {{< var interpolation_all_rmse_cis >}} eV \added{for $\DeltaEST[DSP]{CIS}$} ([@fig-interpolation]). All calculations have been performed on a MacBook Air laptop computer with an M2 processor. We further integrated [coulson]{.smallcaps} with [PySCF]{.smallcaps}[@sunRecentDevelopmentsPySCF2020] to provide alternative algorithms for the SCF convergence and CIS and found the results consistent within numerical accuracy. \added{To give an indication of the computational cost, the SCF calculation for azaphenalene (13 heavy atoms, 14 electrons), takes} {{< var timing_mean_scf_only >}}$\pm${{< var timing_std_scf_only >}} \added{ms (average and standard deviation over 100 runs, respectively). On top of this, DSP takes} {{< var timing_mean_part_scf >}}$\pm${{< var timing_std_part_scf >}} \added{ms, and CIS + DSP takes} {{< var timing_mean_part_cis >}}$\pm${{< var timing_std_part_cis >}} \added{ms.} Oscillator strengths were calculated with the dipole length approximation.[@jorgensenTimedependentHartreeFockCalculations1970;@klessingerExcitedStatesPhotochemistry1995;@damourGroundExcitedStateDipole2023] Some compounds exhibit negative triplet excitation energies and in some case even negative singlet excitation energies, which indicates a restricted-unrestricted instability of the ground state wavefunction. We have here taken the pragmatic approach to ignore these issues, while we will highlight below some examples where it occurs and if it has any effect on the overall results. This approach can be partly justified as the reference data also does not include any stability analysis. The results should anyway be indicative of the gap between the lowest singlet and triplet states with open-shell character, regardless of if they are lower in energy than the closed-shell singlet.

We used four separate datasets for this study. The first is a set of azaphenalenes previously studied by some of us, comprising 256 substituted compounds, for which excitation energies were computed with (2nd-order) algebraic-diagrammatic construction, {{< var reference_level >}}/cc-pVDZ, and geometries were optimized with B3LYP/cc-pVDZ\added{.}[@polliceOrganicMoleculesInverted2021] The second is a set \added{of 138} substituted \deleted{138} azaazulenes with {{< var reference_level >}}/cc-pVDZ excitation energies and optimized with the B97-3c composite method.[@nigamArtificialDesignOrganic2024] The third is a set of 16 rationally designed scaffolds, which including substitutions amounts to {{< var global_n_total >}} unique compounds, optimized at the B97-3c level and with excitation energies at the {{< var reference_level >}}/cc-pVDZ level.[@polliceRationalDesignOrganic2023] The fourth is a set 315 non-alternant hydrocarbons including substitutions, divided into three subsets of size 76, 187 and 52, optimized at the $\omega$B97X-D/def2-TZVP level and with excitation energies at the CC2/aug-cc-pVTZ level. \added{Below, we compare our PPP-based method to the reference levels at the DFT-optimized geometries given in the datasets. For the azaphenalene dataset, we also investigated the impact of geometry. An initial geometry of each molecule was generated with the EmbedMolecule function in RDKit,}[@RDKitOpensourceCheminformatics]\added{ and then optimized with MMFF94.}[@halgrenMerckMolecularForce1996]\added{ This geometry was then further refined with either the GFN-FF force field,}[@spicherRobustAtomisticModeling2020]\added{ the GFN2-xTB semi-empirical method,}[@bannwarthGFN2xTBAccurateBroadly2019] \added{or the ANI-1ccx machine learning potential}[@smithApproachingCoupledCluster2019]\added{ (as implemented in TorchANI}[@gaoTorchANIFreeOpen2020]\added{ and using ASE as the optimization backend}[@hjorthlarsenAtomicSimulationEnvironment2017]\added{).}

@@

We first demonstrate that our method is capable of capturing the inverted $\DeltaE{ST}$ for some model compounds. As shown by Toyota and co-workers, pentalene at the ideal  $D_\mathrm{2h}$ geometry has an inverted gap, while relaxation to the minimum with $C_\mathrm{2h}$ symmetry \added{(using B97-3c)} leads to a normal gap.[@kosekiViolationHundMultiplicity1985] In our calculations, pentalene with equal bond lengths of 1.4 Å displays a small HOMO-LUMO overlap of {{< var pentalene_ideal_overlap >}}, leading to a small exchange interaction of only $2K=$ {{< var pentalene_ideal_2_exchange >}} eV (@fig-pentalene-ideal-orbitals). The dynamic spin polarization correction $\DeltaE[DSP]{SCF}$ of {{< var pentalene_ideal_dsp_scf >}} eV leads to a net predicted $\DeltaEST[DSP]{SCF}$ of {{< var pentalene_ideal_t1_s1_dsp_scf >}} eV. Adding additional correlation with CIS leads to a $\DeltaEST[DSP]{CIS}$ of {{< var pentalene_ideal_t1_s1_dsp_cis >}} eV. The simple perturbation theory model allows us to gain additional insight into the contributions to $\DeltaE[DSP]{SCF}$ as they correspond to single excitations from doubly occupied orbitals below the HOMO to unoccupied orbitals above the LUMO (@fig-pentalene-ideal-excitations). Three excitations, HOMO-3 → LUMO+2, HOMO-2 → LUMO+1 and HOMO-1 → LUMO+3 \added{(1 → 7, 2 → 6 and 3 → 8 in }[@fig-pentalene-ideal-excitations]\added{)} are identified as the main contributors to the spin polarization effect, and could potentially be tuned by substituent effects.

![Excitations and their contribution to $\DeltaE[DSP]{SCF}=$ {{< var pentalene_ideal_dsp_scf >}} eV for pentalene at the $D_\mathrm{2h}$ geometry and bond lengths of 1.4 Å.]({{< var fig_pentalene_ideal_excitations >}}.png){#fig-pentalene-ideal-excitations}

Using the relaxed $C_\mathrm{2h}$ geometry leads to dramatic changes in the frontier orbitals, with an increased HOMO-LUMO overlap of {{< var pentalene_overlap >}} and a sizeable exchange interaction of {{< var pentalene_2_exchange >}} eV (@fig-pentalene-orbitals). Consequently, the calculated gap is now normal at $\DeltaEST[DSP]{SCF}=$ {{< var pentalene_t1_s1_dsp_scf >}} eV, aggravated by a diminished $\DeltaE[DSP]{SCF}$ of only {{< var pentalene_dsp_scf >}} eV. A similar result is obtained with CIS: $\DeltaEST[DSP]{CIS}=$ {{< var pentalene_t1_s1_dsp_cis >}} eV. The reference value calculated with {{< var reference_level >}} is {{< var pentalene_t1_s1_ref >}} eV. The excitations contributing to $\DeltaE[DSP]{SCF}$ at the $D_\mathrm{2h}$ geometry have been significantly diminished, only partially alleviated by the addition of a minor stabilization from HOMO-2 → LUMO+3 (\added{2 → 8 in }[@fig-pentalene-excitations]).
